Columnar VCF for Biobank scale datasets using Zarr

### 2024-04-10 paper.tex
Draft of text for subset matrix section

VCF is a text-based format in which each line
describes observations
and metadata for a single variant (usually a position on the genome
where variation occurs in the samples)~\citep{danecek2011variant}. 
Blocks of VCF text are compressed using \texttt{bgzip},
which allows subsets of the file to be retrieved using secondary
indexes stored alongside the original file~\citep{li2011tabix}.
VCF is therefore a \emph{row-wise} format, i.e., where all of the 
data for a record is stored continguously. BCF is similarly 
row-wise, as are the majority of proposed alternative storage formats.
Row-wise storage makes retrieving all information
for a given record straightforward and efficient, and 
allows data to be easily streamed through Unix pipelines (see Discussion).

@@

@@

The data shown in Figure~\ref{fig-whole-matrix-compute} for Zarr and Savvy
is based on custom programs written using their respective APIs
to implement the \texttt{af-dist} operation. The Zarr program uses
the Zarr-Python package to iterate over the decoded chunks of the 
genotype matrix and classifies genotypes within a chunk using a 14 line Python
function, accelerated using the Numba JIT compliler~\cite{lam2015numba}.
The allele frequencies and genotype counts are then analysed to produce 
the final counts within the allele frequency bins with 9 lines of 
Python using NumPy~\cite{harris2020array} functions. Remarkably, this 
short and simple Python program is substantially faster than the 
equivalent compiled C using \texttt{htslib} APIs on BCF (6 hours
vs 20 hours for 1 million samples). The fastest method is the 
C++ program written using the Savvy API. This would largely seem
to be due to Savvy's excellent genotype decoding performance
(up to 5.7GiB/s vs 1.2GiB/s for Zarr on this dataset;
Figure~\ref{fig-whole-matrix-decode}).

\subsection{Subsetting the genotype matrix}
Existing work overwhelmingly emphasises accessibility by variant.
We demonstrate that sgkit/zarr is good for getting subsets
in terms of both variants and samples via afdist on various subsets.
Refer to fig~\ref{fig-subset-matrix-compute}.

\begin{figure}
\includegraphics{figures/subset-matrix-compute}
\caption{Compute performance on subsets of the matrix.
Total CPU time required to run the af-dist calculation for
a subset of 10 samples and 10000 variants from the middle of the matrix
for the data in Figure~\ref{fig-data-storage}.
[TODO Note that in this case the bcftools pipeline requires
an additional step to compute AF etc.]
\label{fig-subset-matrix-compute}}
\end{figure}

@@

\end{itemize}

>>>>

VCF is a text-based format in which each line
describes observations
and metadata for a single variant (usually a position on the genome
where variation occurs in the samples)~\citep{danecek2011variant}. 
VCF is therefore a \emph{row-wise} format, i.e., where all of the 
data for a record is stored continguously. BCF is similarly 
row-wise, as are the majority of proposed alternative storage formats.
Row-wise storage makes retrieving all information
for a given record straightforward and efficient, and 
allows data to be easily streamed through Unix pipelines (see Discussion).

@@

[Short note here on why we choose these tools to
benchmark against, and why there's no point in 
a major benchmarkathon. Some recent papers 
have good benchmarks which the interested 
reader can cross-reference]

@@

The data shown in Figure~\ref{fig-whole-matrix-compute} for Zarr and Savvy
is based on custom programs written using their respective APIs
to implement the \texttt{af-dist} operation. The Zarr program uses
the Zarr-Python package to iterate over the decoded chunks of the 
genotype matrix and classifies genotypes within a chunk using a 14 line Python
function, accelerated using the Numba JIT compliler~\cite{lam2015numba}.
The allele frequencies and genotype counts are then analysed to produce 
the final counts within the allele frequency bins with 9 lines of 
Python using NumPy~\cite{harris2020array} functions. Remarkably, this 
short and simple Python program is substantially faster than the 
equivalent compiled C using \texttt{htslib} APIs on BCF (6 hours
vs 20 hours for 1 million samples). The fastest method is the 
C++ program written using the Savvy API. This would largely seem
to be due to Savvy's excellent genotype decoding performance
(up to 5.7GiB/s vs 1.2GiB/s for Zarr on this dataset;
Figure~\ref{fig-whole-matrix-decode}). Different choices 
of compressor and filters (zstandard with bit shuffle, in this case) 
would likely lead to improved decoding performance in Zarr,
at the expense of somewhat larger files.

\subsection{Subsetting the genotype matrix}
As datasets grow ever larger, the ability to efficiently access subsets 
of the data becomes more and more important. VCF/BCF achieve efficient 
access to the data for genomic ranges 
by compressing blocks of adjacent records using \texttt{bgzip},
and storing secondary indexes alongside the original 
files with a conventional suffix~\citep{li2011tabix}. 
Thus, for a given range query we 
decompress only the necessary blocks and can quickly access
the required records. 
Methods like Savvy and XSI largely follow the same principles [TODO check and 
be more concrete]. The row-wise nature of these methods, however, mean 
that we cannot efficiently subset \emph{by sample}. In the extreme
case, if we want to access only the genotypes for a single sample
we must still read in and decompress the entire dataset.

\begin{figure}
\includegraphics{figures/subset-matrix-compute}
\caption{Compute performance on subsets of the matrix.
Total CPU time required to run the af-dist calculation for
a subset of 10000 variants $\times$ 10 samples from the middle of the matrix
for the data in Figure~\ref{fig-data-storage}.
The \texttt{bcftools} and \texttt{bcftools} pipelines involve
multiple commands required to correctly calculate the AF INFO tag
required by \texttt{bcftools +af-dist}. See the Methods for full details
on the steps performed.
\label{fig-subset-matrix-compute}}
\end{figure}

We illustrate the cost of row-wise encoding in
Figure~\ref{fig-subset-matrix-compute}, where we run the af-dist calculation
on a small fixed-size subset of the genotype matrices of
Figure~\ref{fig-data-storage}. The two-dimensional chunking of Zarr
means that this small subset of the overall matrix can be efficiently
extracted, and therefore the execution time depends very weakly on 
the overall dataset size, with the overall computation requiring around
1 second for 1 million samples. CPU time, for this simple calculation
on a fixed size sub-matrix scales, with the dataset size.
Figure~\ref{fig-subset-matrix-compute-supplemental} shows performance
for the same operation when selecting half of the samples in the 
dataset.

@@

\end{itemize}
[Random fragment to include somewhere]
Row-wise storage of the genotype matrix makes accessing 
sample haplotypes (i.e. the columns),
required for many
applications~\cite[e.g.][]{durbin2014efficient,kelleher2019inferring} 
[TODO add phasing, imputation]
inefficient. The simple approach of storing the genotype matrix
in rectangular chunks allows efficient access both by row \emph{and}
column. Similarly, these rectangular chunks capture patterns
both of similarity between samples and linkage disequilibrium~\cite{mcvean2019linkage} 
between sites, such that conventional compression schemes 
perform very well~\ref{fig-data-storage}.
